### 2012-07-19 Doc/VRIPHYS2012/vriphys2012.tex
reorganizing sections 2 and 3, adding more stuff to sec. 2
git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@429 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

Methods based on continuum mechanics and namely theory of elasticity gained on popularity in computer graphics and real-time simulators ever sinceTerzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.  %% Do we
While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling \footnote{TODO: do we need ref. here?  yes but for medical applications} of deformations, they are still frowned upon in the area of thin structures. 
Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
This theory can be integrated using Finite Elements, but there is two drawbacks: (i) the continuum equations requires square spatial derivations that can not be handled with linear interpolation and (ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).

The usual argument against non-linear FEM based methods is that they are computationally too expensive. 
This is indeed true if lots of elements are used to discretize the object.
A classical strategy, is to use high-polygonal mesh for rendering or collision detection and response, while maintaing a reasonably low number of elements for the mechanics.
The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynoms.
This type of interpolation is necessary because bending continuum model (based on Kirchhoff-Love theory for thin plates)  requires the computation of square spatial derivatives.
In \cite{Comas2010c}, Comas and al.  use an hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
The element is piloted by 3 nodes on which 6 degrees of freedom are defined. 
From this point of view, our work is derived from this approach but we address several noted drawbacks: the formulation misses one rotational degree of freedom, the cubic interpolation is not symmetric and produces $C^0$  discontinuities between elements on edges.
Ubach and al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
So the stiffness forces computed on each element not only depends on nodes of the element but also on all the adjacent nodes. 
This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.

hell element based on the
% formulation of Bézier triangle.
%
% TODO: paper structure
TODO: paper structure

@@

\section{Preliminaries} % {{{

@@

\subsection{Bézier Triangle} % {{{

Bézier triangles were designed in sixties by French mathematician de
Casteljau. As with other Bézier or B-spline surfaces the Bézier triangle is
defined by a mesh of control points that do not necessarily lie on the
surface. The general $n$-th order Bézier triangle requires $(n + 1)(n +
2)/2$ control points and it's surface is defined as:

In our work we use cubic Bézier triangle ($n=3$). Triangle of smaller order
would fail to describe all possible deformations of the element and
triangles of higher order cannot be described unambiguously with only 18
DOFs available to us without adding further constraints on the control
mesh. Cubic triangle is described by 10 control points and the surface is
explicitly defined as (see fig. \ref{fig-bezier}):

\Figure{0.8\linewidth}{bezier}
{Cubic Bézier triangle with mesh of 10 control points.}

\section{Bézier Shell Element} % {{{

The Bézier triangles are defined by a mesh of control points, however the
simulated object is usually described by a triangular mesh. We construct
the Bézier mesh by taking into account the curvature of the triangular
mesh. When constructing the mesh a special care has to be taken to maintain
the continuity across the nodes and edges. During the initialization phase
we partially employ the method described in \cite{Ubach2010} to maintain
$C^0$ continuity on the edges. Each of the control points on the edge is
computed as the intersection of:

We do not repeat this procedure during the simulation. For faster update of
the control points we rigidly attach the edge points to corner points. We
remember the position and orientation of the edge point relative to the
nearest corner point and use this information together with the rotation on
the corner points to reconstruct new position of control points. See Figure
\ref{fig-segments} for the correspondence between edge points and
associated corner points.

$C^1$ continuity on the edges can also be maintained if special care is taken
when computing the position of the central point (see \cite{Ubach2010}). For
simplicity we compute the central point followingly:

% TODO - nodes + DOFs (local + global position)

@@

Our definition of the displacement field $u$ is based on the formulation of the
Bézier triangle \eqref{eq-cubicbez}:

\begin{equation}\label{eq-bezU}
\begin{split}
  u & = U_1 ( \xi_1^3 + 3 \xi_1^2 \xi_2 + 3 \xi_1^2 \xi_3
        + 2 \xi_1 \xi_2 \xi_3 ) \\
    & + U_2 ( \xi_2^3 + 3 \xi_1 \xi_2^2 + 3 \xi_2^2 \xi_3
        + 2 \xi_1 \xi_2 \xi_3 ) \\
    & + U_3 ( \xi_3^3 + 3 \xi_1 \xi_3^2 + 3 \xi_2 \xi_3^3
        + 2 \xi_1 \xi_2 \xi_3 ) \\
    & + 3 \xi_1^2 \xi_2 (P_{4-1} \times \theta_1) \\
    & + 3 \xi_1^2 \xi_3 (P_{5-1} \times \theta_1) \\
    & + 3 \xi_1 \xi_2^2 (P_{7-2} \times \theta_2) \\
    & + 3 \xi_2^2 \xi_3 (P_{6-2} \times \theta_2) \\
    & + 3 \xi_1 \xi_3^2 (P_{8-3} \times \theta_3) \\
    & + 3 \xi_2 \xi_3^2 (P_{9-3} \times \theta_3) \\
    & + 2 \xi_1 \xi_2 \xi_3 (P_{10-1} \times \theta_1 +
        P_{10-2} \times \theta_2 +
        P_{10-3} \times \theta_3)
\end{split}
\end{equation}

\noindent
where $U_1, U_2, U_3$ are the translational displacment vectors of the
corner nodes of the element, $\theta_1, \theta_2, \theta_3$ are the angular
displacements of the corner nodes and $P_{i-j}$ are the segments from
corner nodes to other nodes of the control mesh:

\begin{equation}
    P_{i-j} = P_j-P_i %, \quad i \in [4;10], j \in [1;3]
\end{equation}

@@

We use the fact that the geometry of the shell is based on the formulation
of cubic Bézier triangle \eqref{eq-cubicbez}. For every vertex of the high
resolution mesh we first find the corresponding triangle on the mechanical
mesh that is closest to the vertex and assign barycentric coordinates on
the triangle to this vertex. After every step of the simulation the high
resolution mesh is updated using the assigned barycentric coordinates and
the function of the surface \eqref{eq-cubicbez}.

@@

In next two subsections we present tests to validate the convergence of our
Bézier shell element. The results are compared with the shell element
composed of DKT element and optimal ANDES element \cite{Felippa2003}. The
last subsection presents evaluation of the computational complexity of our
solution.

@@

\Figure{\linewidth}{roof}
{Convergence for modified Scordelis-Lo roof test of DKT+ANDES element and
our Bézier shell element.}

@@

\Figure{\linewidth}{hemisphere}
{Convergence for test on hemisphere with 18° hole for the DKT+ANDES element and
our Bézier shell element.}

@@

On the other hand the choices provide a good trade-off between continuity
and simplicity. Only $ 3 \times 6~\mathrm{DOFs} = 18~\mathrm{DOFs}$ are
necessary while the Bézier triangle has theoretically $ 10 \times
3~\mathrm{DOFs} = 30~\mathrm{DOFs}$. For complex elements with high number
of nodes the interconnection between nodes increases, especially on the
vertices where the size of nodes involved depends also on number of
neighbouring elements. By keeping the element simple the sparsity of
stiffness matrix is not violated.

@@

This paper presents new shell element with geometry and displacement
function based on the formulation of Bézier triangle. It fixes problems
of the previous model \cite{Comas2010c}, namely:

@@

Because of the formulation of Bézier triangle that is intrinsically present
there is a clean and obvious way of mapping high-resolution mesh for
better visualisation. We also provide a mapping of forces and velocities
between simulation mesh and mechanical mesh for collision detection/response
and interaction with the mesh which are essential component for the surgical
simulator.

@@

% vim: fdm=marker tw=75

>>>>

Methods based on continuum mechanics and namely theory of elasticity gained on popularity in computer graphics and real-time simulators ever since Terzopoulos \cite{Terzopoulos1987} presented his work on elastic deformation modeling.  %% Do we
While the methods using finite elements and theory of elasticity are fairly popular in volumetric modeling \footnote{TODO: do we need ref. here?  yes but for medical applications} of deformations, they are still frowned upon in the area of thin structures. 
Yet, a special field of continuum mechanics has studied the deformations of these objects, and has proposed models based on shell theory.
This theory can be integrated using Finite Elements, but there are two drawbacks: (i) the continuum equations requires second spatial derivations that can not be handled with linear interpolation and (ii) the final system of equations that describes the mechanics is non-linear (if the deformations are not small).

The usual argument against non-linear FEM based methods is that they are computationally too expensive. 
This is indeed true if lots of elements are used to discretize the object.
A classical strategy, is to use high-polygonal mesh for rendering or collision detection and response, while maintaing a reasonably low number of elements for the mechanics.
The work of Bouthors et al.\cite{Bouthors2007} is a good application example of this strategy for surface deformations.
But this strategy is not sufficient in our case because the problem of performance is also greatly impacted when using elements that have an interpolation with high order polynoms.
This type of interpolation is necessary because bending continuum model
(based on Kirchhoff-Love theory for thin plates)  requires the computation of second spatial derivatives.
In \cite{Comas2010c}, Comas and al.  use an hybrid interpolation: linear interpolation is used for in-plane deformations and cubic polynomials for bending deformations.
The element is piloted by 3 nodes on which 6 degrees of freedom are defined. 
From this point of view, our work is derived from this approach but we address several noted drawbacks: the formulation misses one rotational degree of freedom, the cubic interpolation is not symmetric and produces $C^0$  discontinuities between elements on edges.
Ubach and al. \cite{Ubach2010}  present a shell element based on B\'ezier triangles that allows for a better behavior.
However, the B\'ezier interpolation defined on each triangle element is based on the motion computed on adjacent triangles.
So the stiffness forces computed on each element not only depends on nodes of the element but also on all the adjacent nodes. 
This greatly reduces the sparsity of the global stiffness matrix and consequently increases the computation time.

% TODO: paper structure
TODO: paper structure

@@

\section{Element Geometry} % {{{

\subsection{B\'ezier Triangle} % {{{

B\'ezier triangles were designed in sixties by French mathematician de
Casteljau. As with other B\'ezier or B-spline surfaces the B\'ezier triangle is
defined by a mesh of control points that do not necessarily lie on the
surface. The general $n$-th order B\'ezier triangle requires $(n + 1)(n +
2)/2$ control points and it's surface is defined as:

In our work we use cubic B\'ezier triangle ($n=3$). Triangle of smaller order
would fail to describe all possible deformations of the element and
triangles of higher order cannot be described unambiguously with only 18
DOFs available to us without adding further constraints on the control
mesh. Cubic triangle is described by 10 control points and the surface is
explicitly defined as (see fig. \ref{fig-bezier}):

\Figure{0.8\linewidth}{bezier}
{Cubic B\'ezier triangle with mesh of 10 control points.}

The B\'ezier triangles are defined by a mesh of control points, however the
simulated object is usually described by a triangular mesh with 6 degrees of freedom per node (position \& rotation).
We construct
the initial B\'ezier mesh by taking into account the curvature of the triangular
mesh. Note that special care has to be taken to maintain
the continuity across the nodes and edges. During the initialization phase
we partially employ the method described in \cite{Ubach2010} to maintain
$C^0$ continuity on the edges. Each of the control points on the edge is
computed as the intersection of:

We do not repeat this procedure during the simulation. For fast update of
the control points we rigidly attach each edge point to nearest corner point (see Figure
\ref{fig-segments} for the correspondence between edge points and
associated corner points).

In the initialization phase we remember the position of edge point relative to the position and orientation of the corner node:

\begin{equation}
    \overline{P_{i-j}} = (R_i^0)^T P_{i-j}^0
\end{equation}

\noindent
where $P_{i-j}^0 = P_j^0 - P_i^0$, with $j \in [4;10]$ being index of the edge node and $i \in [1;3]$ the index of associated corner node, is the segment at the beginning of the simulation and $R_i^0$ the orientation of node $P_i$.

In each step of the simulation we recompute the new position based on 
current orientation of the corner node:

\begin{equation}
    P_j = P_i + P_{i-j} = P_i + R_i \overline{P_{i-j}}
\end{equation}

\noindent
where $P_i$ is the current position of corner node and $R_i$ its orientation.

$C^1$ continuity on the edges can also be maintained if special care is taken
when computing the position of the central point (see \cite{Ubach2010}). For
simplicity we compute the central point like this:

\subsection{Deflection function} % {{{

% TODO: we need to justify how we decide to move the control points

We recall the equation for central node \eqref{eq-central} and
considering how are the edge points of control mesh connected to corner points with
segments $P_{i-j}$:

\begin{equation}
    P_{i-j} = P_j - P_i, \quad j \in [4;9], i \in [1;3].
\end{equation}

We now add small variation to the position ($U_1, U_2, U_3$) and orientation ($\theta_1, \theta_2, \theta_3$) of corner nodes to the equation of the element \eqref{eq-cubicbez}:

\begin{equation}
    \begin{split}
        & n_1 (P_1 + U_1)
        + n_2 (P_2 + U_2)
        + n_3 (P_3 + U_3) \\
        & + n_4 (P_1 + U_1 + P_{1-4} + P_{1-4} \times \theta_1)
        + n_5 (P_1 + U_1 + P_{1-5} + P_{1-5} \times \theta_1) \\
        & + n_6 (P_2 + U_2 + P_{2-6} + P_{2-6} \times \theta_2)
        + n_7(P_2 + U_2 + P_{2-7} + P_{2-7} \times \theta_2) \\
        & + n_8 (P_3 + U_3 + P_{3-8} + P_{3-8} \times \theta_3)
        + n_9 (P_3 + U_3 + P_{3-9} + P_{3-9} \times \theta_3) \\
        & + n_{10} \frac{1}{3} \left(
        \sum_{j=4}^9 (P_i + U_i + P_{i-j} + P_{i-j} \times \theta_i) - \sum_{i=1}^3 (P_i + U_i)
        \right) \\
%
        & = 
    \sum_{i=1}^{10} n_i P_i
      + n_1 U_1
      + n_2 U_2
      + n_3 U_3 \\
      & + n_4 (U_1 + P_{1-4} \times \theta_1)
      + n_5 (U_1 + P_{1-5} \times \theta_1)
      + n_6 (U_2 + P_{2-6} \times \theta_2) \\
      & + n_7 (U_2 + P_{2-7} \times \theta_2)
      + n_8 (U_3 + P_{3-8} \times \theta_3)
      + n_9 (U_3 + P_{3-9} \times \theta_3) \\
      & + n_{10} \frac{1}{3} (U_1 + U_2 + U_3 + \sum_{j=4}^9 P_{i-j} \times \theta_i) \\
%
      & = T + U_1 (n_1 + n_4 + n_5 + \frac{1}{3} n_{10}) \\
      & + U_2 (n_2 + n_6 + n_7 + \frac{1}{3} n_{10}) \\
      & + U_3 (n_3 + n_8 + n_9 + \frac{1}{3} n_{10}) \\
      & + \sum_{j=4}^9 n_j P_{i-j} \times \theta_i
      + n_{10} \frac{1}{3} \sum_{j=4}^9 P_{i-j} \times \theta_i \\
%
      & = T + u
  \end{split}
\end{equation}

\noindent
thus computing the function of deflection $u$.

We have simplified the notation by leaving the variable $i$ free in the above sums involving $P_{i-j}$.
Note however that the expressions are sound and unambiguous because there is only one possible $i$ for each $j$ depending on the connection of segments to the corner nodes (see fig. \ref{fig-segments}).

\section{...} % {{{

@@

@@

...

@@

We use the fact that the geometry of the shell is based on the formulation
of cubic B\'ezier triangle \eqref{eq-cubicbez}. For every vertex of the high
resolution mesh we first find the corresponding triangle on the mechanical
mesh that is closest to the vertex and assign barycentric coordinates on
the triangle to this vertex. After every step of the simulation the high
resolution mesh is updated using the assigned barycentric coordinates and
the function of the surface \eqref{eq-cubicbez}.

@@

In next two subsections we present tests to validate the convergence of our
B\'ezier shell element. The results are compared with the shell element
composed of DKT element and optimal ANDES element \cite{Felippa2003}. The
last subsection presents evaluation of the computational complexity of our
solution.

@@

\Figure{\linewidth}{roof}
{Convergence for modified Scordelis-Lo roof test of DKT+ANDES element and
our B\'ezier shell element.}

@@

\Figure{\linewidth}{hemisphere}
{Convergence for test on hemisphere with 18° hole for the DKT+ANDES element and
our B\'ezier shell element.}

@@

On the other hand the choices provide a good trade-off between continuity
and simplicity. Only $ 3 \times 6~\mathrm{DOFs} = 18~\mathrm{DOFs}$ are
necessary while the B\'ezier triangle has theoretically $ 10 \times
3~\mathrm{DOFs} = 30~\mathrm{DOFs}$. For complex elements with high number
of nodes the interconnection between nodes increases, especially on the
vertices where the size of nodes involved depends also on number of
neighbouring elements. By keeping the element simple the sparsity of
stiffness matrix is not violated.

@@

This paper presents new shell element with geometry and displacement
function based on the formulation of B\'ezier triangle. It fixes problems
of the previous model \cite{Comas2010c}, namely:

@@

Because of the formulation of B\'ezier triangle that is intrinsically present
there is a clean and obvious way of mapping high-resolution mesh for
better visualisation. We also provide a mapping of forces and velocities
between simulation mesh and mechanical mesh for collision detection/response
and interaction with the mesh which are essential component for the surgical
simulator.

@@

% vim: fdm=marker
